---
title: "Extinct Plant Species and their Root Causes"
author: "Jocelyn Martinez Rico"
date: "5/9/2022"
output: 
   prettydoc::html_pretty:
    theme: architect
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Project description: An analysis on the causes for extinction of plant species from before the 1900s to 2020 in Madagascar and Mexico. 

Data taken from the [International Union for Conservation of Nature (IUCN) Red list of Threatened Species](https://www.iucnredlist.org/) and curated by [Florent Lavergne](https://www.behance.net/florentlavergne) for Tidy Tuesday [2020-08-18](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-08-18/readme.md). 

# Load libraries
```{r}
library(here)
library(tidyverse)
library(dplyr)
library(flair)
library(rcartocolor)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(showtext)
library(prettydoc)
```

# Load data
```{r}
plant_data <- read_csv(here("plant_proj/data","plants.csv"))
action_data <- read_csv(here("plant_proj/data","actions.csv"))
threats_data <- read_csv(here("plant_proj/data","threats.csv"))
```

# Data wrangling

```{r}
#remove zero values
clean_threat <- filter(threats_data, threatened > 0)
clean_action <- filter(action_data, action_taken > 0)
```

# Glimpse data
```{r}
head(clean_action)
head(clean_threat)
```

# Extinct Plants from 1920-1939 in North America

```{r}
north <- clean_threat %>%
  filter(continent == "North America", #selects species in North America
         year_last_seen == "1920-1939", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) %>% #selects extinct plants
  slice_max(binomial_name, n = 25) #selects 25 rows
```


```{r, fig.align='center', fig.width=5, fig.height=5}
north %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

## Actions taken 

```{r}
actions1 <- clean_action %>%
  filter(continent == "North America", #selects plants in North America
         year_last_seen == "1920-1939", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) %>% #selects extinct plants in range
  slice_max(binomial_name, n = 25) #selects 25 rows
```

```{r}
actions1 %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```


# Extinct Plants from 2000-2020 in North America
```{r}
two_north <- clean_threat %>%
  filter(continent == "North America", #selects plants in North America
         year_last_seen == "2000-2020", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) #selects extinct plants
```

```{r, fig.align='center', fig.width=5, fig.height=5}
two_north %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

## Actions taken

```{r}
actions2 <- clean_action %>%
  filter(continent == "North America", #selects plants in North America
         year_last_seen == "2000-2020", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) #selects extinct plants
```

```{r}
actions2 %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

# Extinct Plants from 1920-1939 in South America

```{r}
south <- clean_threat %>%
  filter(continent == "South America", #selects plants in South America
         year_last_seen == "1920-1939", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) %>% #selects extinct plants
  slice_max(binomial_name, n = 25) #selects 25 rows
```

```{r}
south %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

## Actions taken 
```{r}
actions3 <- clean_action %>%
  filter(continent == "South America", #selects plants in North America
         year_last_seen == "2000-2020", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) #selects extinct plants
```

```{r}
actions3 %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

# Extinct Plants from 2000-2020 in South America

```{r}
two_south <- clean_threat %>%
  filter(continent == "South America", #selects plants in North America
         year_last_seen == "2000-2020", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) #selects extinct plants
```

```{r}
two_south %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

## Actions taken

```{r}
actions4 <- clean_action %>%
  filter(continent == "South America", #selects plants in North America
         year_last_seen == "2000-2020", #year range
         red_list_category %in% c("Extinct","Extinct in the Wild")) #selects extinct plants
```

```{r}
actions4 %>%
  kbl() %>% #base code for table
  kable_classic(full_width = F, html_font = "Cambria") #adds theme and font
```

# Threat type analysis

```{r}
threat_count <- clean_threat %>%
  select(threat_type, group) #selects type counts
```

```{r, echo=FALSE}
font_add(family = "baskerville", regular = "Baskerville.ttc")
showtext_auto()
```

```{r}
threat_count %>%
  ggplot(aes(x = threat_type, #x axis
             y = group, #y axis
             fill = threat_type, #point fill
             color = threat_type))+ #color fill
  geom_point(size = 4)+ #changes point size
  labs(x = "Threat Type", #x axis
       y = "Type of Plant", #y axis
       fill = "Threat Type:")+ #fills points and renames legend
  theme_light()+ #b/w theme
  theme(text = element_text(family = "baskerville"), #adds font
        axis.text.x = element_text(angle = 45, hjust = 1))+ #angles axis labels
  scale_color_carto_d(name = "Threat Type:", palette = "Vivid") #color to fill
```

# Action type analysis

```{r}
action_count <- clean_action %>%
  select(action_type, group) #selects type counts
```

```{r}
action_count %>%
  ggplot(aes(x = action_type, #x axis
             y = group, #y axis
             fill = action_type, #point fill 
             color = action_type))+ #color fill
  geom_point(size = 4)+ #changes point size
  labs(x = "Action Type", #x axis
       y = "Type of Plant", #y axis
       fill = "Action Type:")+ #fills points and renames legend
  theme_light()+ #b/w theme
  theme(text = element_text(family = "baskerville"), #adds font
        axis.text.x = element_text(angle = 45, hjust = 1))+ #angles axis labels
  scale_color_carto_d(name = "Action Type:", palette = "Antique") #color to fill
```



